Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-12 paper.tex
ENH:  Adding more stuff.

@@

\graphicspath{{./Figures/}
                          }

@@

\def\journal{Neuroscience}%%% write here for which journal %%%
\def\DOI{}
\def\articleType{Methods}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Tustison {et~al.}} %use et al only if is more than 1 author
\def\Authors{Nicholas J. Tustison\,$^{1,*}$ and Brian A. Avants\,$^{2}$
 }
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$University of Virginia, Department of Radiology and Medical Imaging, Charlottesville, VA, USA \\
$^{2}$Penn Image Computing and Science Laboratory, University of Pennsylvania, Department of Radiology, Philadelphia, PA, USA }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Nick Tustison}
\def\corrAddress{University of Virginia, Department of Radiology and Medical Imaging,  480 Ray C Hunt Drive, Charlottesville, VA, 22903}
\def\corrEmail{ntustison@virginia.edu}

@@

\title[Explicit B-spline Regularization]{Explicit B-spline Regularization in Diffeomorphic Image Registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several 
advantages including ease of algorithmic understanding, 
good performance, desirable geometric properties, and several available 
implementations.  Current research was preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.  
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

Parallel to this branch of algorithmic progress are the informally 
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are detailed in 
the works of \cite{bro-nielsen1996} and \cite{pennec1999} who observe that
Gaussian smoothing, characteristic of Demons, of the update or displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation, respectively.  However, the use of such 
approximations entails that physical properties, such as topological
regularity, are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach 
where, rather than produced by Gaussian convolution, regularized fields are produced using B-splines for roughly approximating the solution to the viscous fluid system.  This was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.  As mentioned in \cite{tustison2009}, this preconditioning is used to reduce the hemstitching induced by the traditional gradient-based FFD formulation.
Further details can be gleaned from the ITK implementation of this 
earlier work%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'') 
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998}, 
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm  
was proposed in \cite{Beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  
Symmetrical considerations in the velocity field parameterization, for example, are discussed in \cite{avants2008} in the context of a cross correlation similarity metric which minimizes the bias of the resulting transformation in selecting the identity of the fixed and moving images.  A greedy version of this algorithm which has proven successful in general application \citep{Klein2009}.

@@

Alternatively, the FFD 
variant reported in \cite{rueckert2006} enforced diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.  Another FFD registration
incorporated the recent log-Euclidean framework for enforcing diffeomorphic
transformations \cite{Modat2011}.
Recently, the work of 
\cite{de-craene2011} combined these registration concepts into a single
framework called {\em temporal free-form deformation} in which the 
time-varying velocity field characteristic of LDDMM-style algorithms
is modeled using a 4-D B-spline object (3-D + time).  Integration of 
the velocity field yields the mapping between parameterized time points.

%
%
%framework developed independently of Demons and LDDMM/DARTEL.
%But then the FFD people saw how useful the diff. properties were and
%tried to integrate B-splines into the diff. framework.
%
%
%The early deformable work of 
%\begin{itemize}
%  \item xx Early deformable work \cite{broit1981,bajcsy1989}.
%  \item FFD \cite{Sleziski,Rueckert1989}
%  \item Christensen's fluid work
%  \item Optical flow?
%  \item Fast fluid registration of medical images (Bro-Nielsen and Gramkow).
%  \item Consistent image registration
%  \item Demons and Diff demons
%  \item LDDMM 
%  \item DARTEL
%  \item Arno Klein's study
%\end{itemize}
%
%
%
%Early work included the elastic deformation

%Significant algorithmic developments characterizing modern intensity-based
%image registration research include the B-spline parameterized approach
%(so called {\em free-form deformation}) with early contributions 
%including \cite{szeliski1997,thevenaz1998,rueckert1999}.  Amongst the 
%variant extensions, the 
%{\em directly manipulated free-form deformation} approach  \cite{tustison2009} 
% addressed the hemstitching issue associated with steepest descent traversal of
%problematic energy topographies during the course of optimization.
%%caused by the distribution of the uniform 
%%B-spline shape functions over the transformation domain 
%
%
%In this work, we describe our extension to these methods.  Similar to 
%\cite{de-craene2011}, we also use an $N$-D + time B-spline object to 
%represent the characteristic velocity field.  However, we use the 
%directly manipulated free-form deformation optimization formulation to improve 
%convergence during the course of optimization.  This also facilitates
%modeling temporal periodicity and
%the enforcement of stationary boundaries consistent with diffeomorphic
%transforms.
%We also incorporate B-spline mesh multi-resolution capabilities
%for increased control during registration progression.  
%Most importantly, we also describe
%the parallelized algorithmic implementation as open source available through the Insight Toolkit.%
%\footnote{
%http://www.itk.org/
%}
%
%We first describe the methodology by laying out a mathematical description 
%of the various algorithmic elements coupled with implementation details
%where appropriate.  This is followed by an evaluation on publicly available
%brain data.

\section{Material and Methods}
%Provide sufficient detail to allow the work to be reproduced. Methods already published should be indicated by a reference: only relevant modifications should be described.

Three popular algorithms [1,2,3] (and their B-spline analogs) were implemented as part of the recent refactoring of the Insight Toolkit (ITK) [5] and included as part of the Advanced Normalization Tools (ANTs) repository [6].  Thus, all code used for this work is available as open source.   For comparison, the ITK-based, well-vetted SyN algorithm [3], which uses explicit Gaussian regularization, was used to provide the canonical mappings in the MICCAI 2012 multi-atlas labeling challenge [7] between 20 training subjects and 15 testing subjects (140 labeled structures).  Using the exact same parameters, except those specific to the regularization, we produced the analogous B-spline SyN mappings for the same data and compared transformations using the Dice overlap.

\subsection{Public Data}

@@

@@

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\graphicspath{{./Figures/}
              {./Results/}}

@@

\def\journal{Neuroscience}%%% write here for which journal %%%
\def\DOI{}
\def\articleType{Methods}
\def\keyFont{\fontsize{8}{11}\helveticabold }
%\def\firstAuthorLast{Tustison {et~al.}} %use et al only if is more than 1 author
\def\firstAuthorLast{Tustison and Avants} %use et al only if is more than 1 author
\def\Authors{Nicholas J. Tustison\,$^{1,*}$ and Brian B. Avants\,$^{2}$
 }
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$University of Virginia, Department of Radiology and Medical Imaging, Charlottesville, VA, USA \\
$^{2}$Penn Image Computing and Science Laboratory, University of Pennsylvania, Department of Radiology, Philadelphia, PA, USA }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Nick Tustison}
\def\corrAddress{University of Virginia, Department of Radiology and Medical Imaging,  480 Ray C Hunt Drive, Charlottesville, VA, 22903}
\def\corrEmail{ntustison@virginia.edu}

@@

\title[Explicit B-spline regularization]{Explicit B-spline regularization in diffeomorphic image registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity,
good performance, desirable properties, and several available
implementations.  Current research was
preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

Parallel to this branch of algorithmic progress are the informally
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are detailed in
the works of \cite{bro-nielsen1996} and \cite{pennec1999} who observe that
smoothing via Gaussian convolution, a defining characteristic of Demons,
of the update or displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation, respectively.  However, the use of such
approximations entails that physical properties, such as topological
regularity, are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space
of regularized fields using Gaussian convolution, gradient fields
are projected to a smooth space characterized by the B-spline
basis functions.  This analogy was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect induced by the ill-conditioned nature of the traditional gradient-based FFD formulation.
Further details can be gleaned from the ITK implementation of this
earlier work%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric which minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

Although many extensions of LDDMM rely
on some form of Gaussian convolution for
regularization \citep[e.g.][]{risser2011}, there has been significant interest in constraining
FFD approaches to the space of diffeomorphisms.
An early attempt reported in \cite{rueckert2006} enforced
diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.
\cite{modat2011} incorporated the log-Euclidean
framework for enforcing diffeomorphic transformations and ensuring invertibility.  Similarly, the work of \cite{de-craene2011} provided
a full LDDMM-style algorithm based on B-splines called
{\em temporal free-form deformation} in which the
time-varying velocity field
is modeled using a 4-D B-spline object (3-D + time).  Numerical integration of the mapping propagated within the velocity field yields the transform between parameterized time points.

As alluded to earlier, B-spline approximation can also be used
for regularizing vector fields in an analogous fashion as
Gaussian convolution.  In this vein and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to
represent the characteristic velocity fields.  However, we use the
directly manipulated free-form deformation formulation to improve the solution convergence and actual results.  This also facilitates modeling
temporal periodicity and the enforcement of stationary boundaries.

Both this latter work \citep{tustison2012a} and our earlier work
\citep{tustison2009} demonstrate that our DMFFD framework is potentially
applicable to the
entire gamut of diffeomorphic registration algorithms and provides
unique ``flavors'' of smoothing possibilities.  However, differences
between the two approaches produce characteristically different
solutions.  In addition to
smoothing kernel differences, Gaussian
convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD
approach.  Also, whereas
convolution operates entirely within the discrete space,
the fitting process constructs a continuous object prior
to any discrete sampling.  Further elaboration on these differences
are discussed in subsequent sections.

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1TimeVaryingVelocityFieldTransform.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1TimeVaryingBSplineVelocityFieldTransform.html
\end{itemize}
},
DARTEL%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianExponentialDiffeomorphicTransform.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineExponentialDiffeomorphicTransform.html
\end{itemize}
},
 and SyN%
\footnote{
\begin{itemize}
\item http://www.itk.org/Doxygen/html/classitk\_1\_1SyNImageRegistrationMethod.html
\item http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSyNImageRegistrationMethod.html
\end{itemize}
})
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the symmetric normalization (SyN) algorithm, our focus in this work is
its B-spline equivalent which we denote as ``B-spline SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors \citep{tustison2012}).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

\section{Material and Methods}

\subsection{Theoretical Overview}

Given the spatial domain $\Omega$ of $d-$dimensionality
defined over image $I$, a diffeomorphic
mapping, $\phi$, parameterized over $t \in [0,1]$ transforms the image
$I$ to the target image $J$ using $I\circ\phi(\mathbf{x},1)$ where the
geodesic path $\phi(\mathbf{x},t)$ is described by \citep{beg2005}
\begin{align}
  \inf_{\phi} \left( \int_0^1 \|v(t)\|_L^2 dt +
                     \int_{\Omega} | I \circ \phi^{-1}(\mathbf{x},1) - J |^2 d\Omega
              \right).
  \label{eq:lddmm}
\end{align}
$\phi$ is generated as the solution of the ordinary differential equation
\begin{align}
  \frac{d \phi(\mathbf{x},t)}{dt} = v( \phi(\mathbf{x},t), t ),\,\, \phi(\mathbf{x},0) = \mathbf{Id}
\label{eq:ode}
\end{align}
where $v$ is a time-dependent smooth field (as dictated by the functional norm $L$), $v : \Omega \times t
\rightarrow \mathrm{R}^d$ parameterized by $t \in [0,1]$.  Diffeomorphic mappings
between parameterized time points $\{t_a,t_b\} \in [0,1]$
are obtained from  Eq. (\ref{eq:ode}) through integration of the transport
equation, viz.
\begin{align}
  \label{eq:integral}
\phi(\mathbf{x},t_b) &= \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} v(\phi(\mathbf{x}), t) dt.
\end{align}
%and $L$ is the functional norm which enforces velocity field regularity.

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(1-t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}
\end{align}
with extension to arbitrary similarity metric choice, i.e. the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{SyN.pdf}
  \caption{Illustration of the greedy SyN formulation.  Given images
  $I_A$ and $I_B$, the symmetric set-up requires finding the two
  transform pairs $\left(\phi_1,\phi_1^{-1}\right)$
  $\left(\phi_2,\phi_2^{-1}\right)$ which map to/from
  the respective images to the midway point. During
  optimization, the update field at each iteration is
  determined from the metric field gradient taken at
  the midway point, i.e.
  $\nabla \Pi_{\sim} \left(I\circ\phi_1(0.5),J\circ\phi_2(0.5)\right)$.
  The full
  forward and inverse transforms are found through
  composition, i.e. $\phi=\phi_1 \circ \phi_2^{-1}$ and
  $\phi^{-1}=\phi_2 \circ \phi_1^{-1}$.
  }
  \label{fig:syn}
\end{figure}

\subsubsection{St. Nava's Theory of Greed and Original SyN}

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common, contemporary processing infrastructures.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  In this variant, the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  The algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

\begin{algorithm}
\caption{SyN algorithmic overview}
\label{alg:syn}
\begin{algorithmic}
\State $\phi_i \leftarrow \mathbf{Id}$, $\phi_i^{-1} \leftarrow \mathbf{Id}$
\Comment $i \in \{1,2\}$
\State $n \leftarrow 1$
\While {not converged}
  \State $v_1^n \leftarrow \nabla \Pi_{\sim} \left(I\circ\phi_1^{n-1},J\circ\phi_2^{n-1}\right)$
  \State $v_2^n \leftarrow \nabla \Pi_{\sim} \left(J\circ\phi_2^{n-1},I\circ\phi_1^{n-1}\right)$
  \State $v_i^n \leftarrow S_v( v_i^n )$      \Comment $S_v$ is a smoothing operation on the update transform field
  \State $\phi_i^n \leftarrow S_\phi( v_i^n \circ \phi_i^{n-1} )$      \Comment $S_\phi$ is a smoothing operation on the total transform field
  \State $\left(\phi_i^n\right)^{-1} \leftarrow Inv\left(\phi_i^n, \left(\phi_i^{n-1}\right)^{-1}\right)$
    \Comment Inverse field estimation described in \cite{avants2008}
  \State $n \leftarrow n + 1$
\EndWhile
\State \Return $\phi \leftarrow \phi_1 \circ \phi_2^{-1}$, $\phi^{-1} \leftarrow \phi_2 \circ \phi_1^{-1}$
\end{algorithmic}
\end{algorithm}

\subsubsection{Directly Manipulated Free-Form Deformation Diffeomorphic Analogs}

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.
Extending the work presented in \cite{tustison2009,tustison2012},
the DMFFD approach is a viable and practical alternative
for explicit regularization in diffeomorphic image registration.

As we mentioned previously, it has been shown that the preconditioned
gradient of the B-spline object is equivalent to a B-spline approximation
smoothing operation on the spatial derivative of the similarity metric
gradient \citep{tustison2009}. In the case of
$d$-dimensional image registration, the time-dependent velocity field can be
represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field and $B(\cdot)$ are the univariate B-spline
basis functions separately modulating regularity in the solution for each parametric dimension.  The preconditioned gradient analog for updating the velocity
field is as follows \citep{tustison2012}:
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi_\sim$,
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which is a slight modification of Eqn. (21) in \cite{tustison2009}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
$r$ is the spline order in all dimensions%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $c$ indexes the spatio-temporal dense metric gradient sample.
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.

@@

Although various methods exist for solving Eqns. (\ref{eq:ode}) and (\ref{eq:integral}),
we use $4^{th}$-order Runge-Kutta, i.e.
\begin{align}
  \phi_{n+1} &= \phi_{n} + \frac{1}{6}\left( k_1 + 2k_2 + 2k_3 + k_4 \right) \\
  t_{n+1} &= t_{n} + \Delta t \\
  \phi_0 &= \phi(t_0)
\end{align}
where
\begin{align}
  k_1 &= v\left( \phi_{n}, t_{n} \right)\Delta t \\
  k_2 &= v\left( \phi_{n} + \frac{k_1}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_3 &= v\left( \phi_{n} + \frac{k_2}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_4 &= v\left( \phi_{n} + k_3, t_{n} + \Delta t \right)\Delta t
\end{align}
which provides a more stable and reliable alternative than other numerical
methods \citep{press2007}.

\subsection{Evaluation Data}

We take inspiration from the well-known
Klein comparative study in which 14 image registration
algorithms were evaluated based on performance
on publicly available labeled brain data \citep{klein2009}.
Based on the final rankings, the SyN algorithm is one
of the top-performing registration algorithms.  Its relative
performance coupled with its public availability has made
it a de facto standard for algorithmic comparison.

For our evaluation, we used the same data as that in the
Klein study.  Specifically, we used the data sets denoted as:
\begin{itemize}
  \item MGH10
  \item CUMC12
  \item IBSR18
  \item LPBA40
\end{itemize}
which are available for download from Arno Klein's website.%
\footnote{
http://mindboggle.info/papers/evaluation\_NeuroImage2009.php
}
We also use the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} which we denote as MICCAI35.  This T1-weighted
MRI data set consists of 35 subject MRI taken from the
Oasis database%
\footnote{
http://www.oasis-brains.org
}.  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allCUMC.pdf}
  \caption{}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allIBSR.pdf}
  \caption{Currently redoing on the cluster.}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allLPBA.pdf}
  \caption{}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allMGH.pdf}
  \caption{Currently redoing on the cluster.}
\end{figure}

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{allMICCAI.pdf}
  \caption{}
\end{figure}
